\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-05 paper.tex
Add link et al paper
Closes #446

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestry
(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
introductions).
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
grundler2024geographic,
huang2024estimating}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).

%Other subtly different concepts that we must be careful to distinguish
%are the ``true'' ARG, % I don't see a better word for "true" here
%describing the actual history of a sample of genomes,
%and a ``population'' ARG which is
%the true ARG of every individual in a population.
%True ARGs are the underlying real history of a sample,
%perfectly resolved into binary splits and mergers by
%%the cellular processes of meiosis and mitosis, regardless
%of sampling density or population processes
%(Appendix~\ref{sec-cell-lineages-and-args}).
%Although population-scale true ARGs unquestionably exist,
%they can also never be entirely known, in part because of a fundamental
%lack of mutational information.
%Even if mutation rate were high enough to uniquely identify every
%recent branching point,
%such a high rate would saturate the genome with mutations
%and obscure deeper history.

% Population ARGs are here, now.
%Population ARGs may seem fanciful, but the scale of
%modern datasets makes it necessary for us to grapple with the idea.
%The UK Biobank (UKB), for example, has
%genotype data for around 500,000 humans~\citep{bycroft2018genome},
%along with exome~\citep{backman2021exome} and
%whole genome sequence~\citep{halldorsson2022sequences} data
%for large subsets of the cohort.
%UKB is just one of many such population-scale sequencing
%projects~\citep[e.g.][]{turnbull2018hundred,
%karczewski2020mutational,tanjo2021practical}.
%Agricultural datasets are on a similar scale,
%and also include dense multi-generational sampling
%and near-perfect pedigree
%information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
%Recent advances have made it possible to actually
%\emph{estimate} ARGs at population scale:
%ARGs have already been inferred for the 500,000
%humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
%and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
%While this new population-scale reality presents many
%exciting opportunities, it also poses substantial challenges
%to existing methodologies.

% What's the problem?
%A major problem currently facing the field is that
%classical mathematical formalisms and terminology
%cannot adequately describe these vast inferred ARGs.
%Fundamentally, these formalisms assume that an
%ARG is known in complete detail
%and are not suited to describing partial or approximate knowledge.
%As we are actively inferring ARGs at the population scale,
%and such ARGs can never be known in complete detail,
%there is currently a substantial gap between our theoretical frameworks
%and practical application.

The breakthroughs in scale achieved by recent
methods are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history
(Appendix~\ref{sec-survey-arg-infer}).
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced in \noderef{k} and \noderef{p} before \noderef{q} is reached.
Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible in an eARG.
Because the ancestral material is stored with each edge of a gARG, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events.
Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

>>>>

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestry.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
link2023tree,
grundler2024geographic,
huang2024estimating}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
methods are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history
(Appendix~\ref{sec-survey-arg-infer}).
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
introductions)

@@

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced in \noderef{k} and \noderef{p} before \noderef{q} is reached.
Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible in an eARG.
Because the ancestral material is stored with each edge of a gARG, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events.
Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a set of samples.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.
